---
title: "geo-hw-pred-v2"
author: "Kevin Ruiz"
format: html
editor: visual
---

The `cow.rda` file contains a data frame with 200 spatial observations observed throughout the state of Colorado. The data frame includes `easting` and `northing` coordinates (UTM, scaled by 1000 meters) and Tungsten measurements (`W`) at each site (mg/kg). Read in the data as a data frame from the file.

# Preliminaries

Load the **gstat** and **geoR** packages. Load the `cow.rda` file.

```{r}
load('cow.rda')
```

```{r}
library(gstat)
library(geoR)
```

# Problem 0

Create `cow_sf`, an `sf` version of `cow` that uses `easting` and `northing` for `coords`.

Create `geocow`, a `geodata` object from the **geoR** package.

Also run the following code.

```{r}
set.seed(1)

poly_sf <- sf::st_polygon(x = list(cbind(
  c(152, 152, 753, 753, 152),
  c(4100, 4544, 4544, 4100, 4100))))



grid_sf <- sf::st_sample(poly_sf, size = 100,
                         type = "regular")

cow_sf <- sf::st_as_sf(cow,
                       coords = c("easting", "northing"))

geocow = as.geodata(cbind(cow$easting , cow$northing,cow$W))



```

**Solution**

# Problem 1 (Using geoR)

Use REML estimation with a constant mean function to fit the following covariance models to the data:

a.  An isotropic spherical model.
b.  A geometric anisotropic spherical model.
c.  An isotropic exponential model.
d.  A geometric anisotropic exponential model.

Make a table that provides the following information about each model: model, partial sill ($c$), range parameter ($a$ or $a_{min}$), nugget ($c_0$), direction of the major axis in degrees, (if appropriate), anisotropy ratio $a_{max}/a_{min}$, AIC of the fit.

Note: starting values make a HUGE difference for the anisotropic models, so you probably want to experiment a bit.

Which model is the best choice?

```{r}
variog1b = variog(geocow)
plot(variog1b)

###Isotropic Spherical
lfit_sph = likfit(geocow,
                  ini.cov.pars = c(1.5, 500),
                  nugget = 0.5,
                  cov.model = "spherical",
                  lik.method = "REML",
                  trend = "cte")

lfit_sph

v <- variog4(geocow)
plot(v)
amax <- 600
amin <- 300
a <- 600/300


###Anisorpic Spherical
lfit1 = likfit(geocow, ini.cov.pars = c(1.5, 500),
               nugget = 0.5,
               cov.model = "spherical",
               lik.method = "REML",
               psiA = 70/180*pi, 
               psiR = a,
               trend = "cte")


####Isotropic Exp
lfit_exp = likfit(geocow,
                  ini.cov.pars = c(1.5, 500),
                  nugget = 0.5,
                  cov.model = "exponential",
                  lik.method = "REML",
                  trend = "cte")

###Anisorpic Exp
lfit2 = likfit(geocow, ini.cov.pars = c(1.5, 500),
               nugget = 0.5,
               cov.model = "exponential",
               lik.method = "REML",
               psiA = 70/180*pi, 
               psiR = a,
               trend = "cte")

model <- c("Isotropic Spherical", "Anisorpic Spherical", "Isotropic Exp", "Anisorpic Exp")

```
