---
title: "Hospitals Mexico Project"
author: "Kevin Ruiz"
format: html
editor: visual
---

```{r}
library(readr)

Merged_CLUES_2016 <- read_csv("Merged_CLUES_2016.csv")
head(Merged_CLUES_2016)
```

You can add options to executable code like this

```{r}
library(sf)
library(readxl)
library(ggplot2)
library(dplyr)
unzip("mx_shp.zip", exdir = "mx_shp")
mex_shp <- st_read("mx_shp/mx.shp")

Merged_CLUES_2016_clean <- Merged_CLUES_2016 |>
  dplyr::filter(!is.na(longitude), !is.na(latitud)) 

pts_sf <- st_as_sf(Merged_CLUES_2016_clean, coords = c("longitude", "latitud"),  crs = 4326)

plot(st_geometry(mex_shp), col = "white", border = "gray")
plot(st_geometry(pts_sf), col = "red", pch = 5, add = TRUE, cex = .5, size =1)

```

Some exploratory graphs showing where we have a lot of our data, for example in the state of chihuahua we see a lot of our health facilities somewhat clustered in the regions near all of the big cities with little to none rural ones.

```{r}
chih_sf <- mex_shp |> 
  filter(name == "Chihuahua")

chih_pts <- pts_sf |> 
  filter(state_code == 8)


plot(st_geometry(chih_sf), col = "white", border = "gray")
plot(st_geometry(chih_pts), col = "red", add = TRUE, cex = .5, size = 1)

```

Now for us to actually do some of the analysis I went ahead and aggregated the data by state municipality code and year. This can help us use spatial statistic methods used in regional count data which helps simplify our analysis.

```{r}
#lets load in our dataset with the aggregation done. 

data <- read_csv("Recursos_Municipio_Year.csv")
```

With our data loaded in we can start by doing some basic data exploration, first lets look at the
