---
title: "Hospitals Mexico Project"
author: "Kevin Ruiz"
format: html
editor: visual
---

```{r}
library(readr)
library(sf)
library(readxl)
library(ggplot2)
library(dplyr)
library(smerc)

Merged_CLUES_2016 <- read_csv("Merged_CLUES_2016.csv")
head(Merged_CLUES_2016)
```

You can add options to executable code like this

```{r}

unzip("mx_shp.zip", exdir = "mx_shp")
mex_shp <- st_read("mx_shp/mx.shp")

Merged_CLUES_2016_clean <- Merged_CLUES_2016 |>
  dplyr::filter(!is.na(longitude), !is.na(latitud)) 

pts_sf <- st_as_sf(Merged_CLUES_2016_clean, coords = c("longitude", "latitud"),  crs = 4326)

plot(st_geometry(mex_shp), col = "white", border = "gray")
plot(st_geometry(pts_sf), col = "red", pch = 5, add = TRUE, cex = .5, size =1)

```

Some exploratory graphs showing where we have a lot of our data, for example in the state of chihuahua we see a lot of our health facilities somewhat clustered in the regions near all of the big cities with little to none rural ones.

```{r}
chih_sf <- mex_shp |> 
  filter(name == "Chihuahua")

chih_pts <- pts_sf |> 
  filter(state_code == 8)


plot(st_geometry(chih_sf), col = "white", border = "gray")
plot(st_geometry(chih_pts), col = "red", add = TRUE, cex = .5, size = 1)

```

Now for us to actually do some of the analysis I went ahead and aggregated the data by state municipality code and year. This can help us use spatial statistic methods used in regional count data which helps simplify our analysis.

```{r}
#lets load in our dataset with the aggregation done. 

data <- read_csv("Recursos_Municipio_Year.csv")
pop <- read_excel("00_Republica_mexicana/1_Grupo_Quinq_00_RM.xlsx") 

```

With our data loaded in we can start by doing some basic data exploration. First, however, lets get the population by municipality for each year.

```{r}

pop_total_by_gender <- pop %>%
  select(CLAVE, POB_TOTAL, ANO) %>%
  rename(Year = ANO) %>%
  filter(Year > 2000, Year < 2011)

pop_total <- pop_total_by_gender %>%
  group_by(CLAVE, Year) %>%
  summarise(
    POB_TOTAL = sum(POB_TOTAL, na.rm = T),
    .groups = "drop"
  )

data_clave <- data %>%
  rename(CLAVE = State_Municipality)
# Number of hopsitals plus a column for our population total for that year. 
Hosp_pop <- data_clave %>%
  left_join(pop_total, by = c("CLAVE", "Year"))
```

```{r}
## Loading in our shape file for the municipaliteis goetten from government website

mex_shp <- st_read("mun21gw/mun21gw.shp")

# Using SF to get centroids for the shapes
mun_centroids <- st_centroid(mex_shp)

# Getting Centroid coords
coords <- st_coordinates(mun_centroids)

# Making the CLAVE match up as well as extracting our coordinates
mun_centroids <- mun_centroids %>%
  mutate(
    CVE_ENT = as.numeric(CVE_ENT),
    CVE_MUN = as.numeric(CVE_MUN),
    CLAVE = CVE_ENT * 1000 + CVE_MUN,
    lon = coords[,1],
    lat = coords[,2]
  )

# Having our coords 
names(mun_centroids)
mun_xy <- mun_centroids %>%
  st_drop_geometry() %>%
  select(CLAVE, lon, lat)

# This is all of the municiplaites
Hosp_pop_xy <- Hosp_pop %>%
  left_join(mun_xy, by = "CLAVE") 


# With this we are only looking at the municiplaites in the northern part of mexico
north_states <- c("02", "05", "08", "19", "26", "28")
Hosp_pop_north <- Hosp_pop_xy %>%
  left_join(mun_xy, by = "CLAVE") %>%
  mutate(CLAVE = sprintf("%05d", as.integer(CLAVE))) %>%
  mutate(state_code = substr(CLAVE, 1, 2)) %>%
  filter(state_code %in% north_states)



```

This is the Beasnag Nuell Method on the whole population of Mexico, or more specifically in regards to

```{r}
year_of_interest <- 2010

Hosp_pop_xy_2010 <-  Hosp_pop_xy %>% filter(Year == year_of_interest)

coords <- as.matrix(Hosp_pop_xy_2010[, c("lon", "lat")])
cases  <- Hosp_pop_xy_2010$num_of_total
pop    <- Hosp_pop_xy_2010$POB_TOTAL
ex     <- sum(cases) / sum(pop) * pop  

cstar <- 12  # e.g., windows that contain 5 hospitals

bn_out <- bn.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  cstar  = cstar,
  alpha  = 0.05,
  longlat = FALSE,
  modified = FALSE
)

bn_out  


cl_bn <- clusters(bn_out)

str(cl_bn)

cl_bn$location$idx

Hosp_pop_xy_2010$bn_cluster <- "None"

Hosp_pop_xy_2010$bn_cluster[cl_bn$location$idx] <- "High hospital density"

library(ggplot2)

ggplot(Hosp_pop_xy_2010, aes(x = lon, y = lat)) +
  geom_point(aes(color = bn_cluster), size = 1.4, alpha = 0.85) +
  scale_color_manual(values = c(
    "None" = "gray80",
    "High hospital density" = "#d73027"
  )) +
  coord_fixed() +
  labs(
    title = "Besag–Newell Cluster of Hospital Supply (2010)",
    subtitle = paste0("c* = ", cstar, " hospitals"),
    x = "Longitude",
    y = "Latitude",
    color = "BN Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )


```

```{r}
Hosp_pop_xy_2010$expected_hosp <- sum(Hosp_pop_xy_2010$num_of_total) /
                                  sum(Hosp_pop_xy_2010$POB_TOTAL) *
                                  Hosp_pop_xy_2010$POB_TOTAL

Hosp_pop_xy_2010$shortage <- Hosp_pop_xy_2010$expected_hosp -
                             Hosp_pop_xy_2010$num_of_total

Hosp_pop_xy_2010$shortage[Hosp_pop_xy_2010$shortage < 0] <- 0

ggplot(Hosp_pop_xy_2010, aes(x = lon, y = lat)) +
  geom_point(aes(color = shortage), size = 1.5) +
  scale_color_viridis_c(option = "magma") +
  coord_fixed() +
  labs(
    title = "Hospital Shortage Index (2010)",
    subtitle = "Expected − Observed Hospitals",
    x = "Longitude",
    y = "Latitude",
    color = "Shortage"
  ) +
  theme_minimal()

```

```{r}
nstar <- 100000  # total population in each CEPP window (tune this!)

cepp_out <- cepp.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  nstar  = nstar,
  nsim   = 499,
  alpha  = 0.05,
  longlat = FALSE
)

cepp_out
summary(cepp_out)
plot(cepp_out)

```

```{r}
scan_out <- scan.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  nsim   = 499,
  alpha  = 0.05,
  ubpop  = 0.5,       # max fraction of total population allowed in a cluster
  longlat = FALSE,
  type   = "poisson"  # counts with population offset
)

scan_out
plot(scan_out, idx = 1:3)      # first 3 clusters
clusters(scan_out)             # extract cluster membership, RR, etc.

```

```{r}
coords <- as.matrix(df_year[, c("x", "y")])
w      <- dweights(coords, kappa = 1)  # distance-based weights

tango_out <- tango.test(
  cases = cases,
  pop   = pop,
  w     = w,
  nsim  = 499    # Monte Carlo p-value
)

tango_out
plot(tango_out)

```

```{r}
years <- sort(unique(Hosp_pop_xy$Year))

results_scan  <- list()
results_tango <- list()

for (yy in years) {
  this <- Hosp_pop_xy %>% filter(Year == yy)

  coords <- as.matrix(this[, c("lon", "lat")])
  cases  <- this$num_of_total
  pop    <- this$POB_TOTAL

  # scan
  results_scan[[as.character(yy)]] <- scan.test(
    coords = coords,
    cases  = cases,
    pop    = pop,
    nsim   = 199,
    alpha  = 0.05,
    ubpop  = 0.5
  )

  # tango
  w <- dweights(coords, kappa = 1)
  results_tango[[as.character(yy)]] <- tango.test(
    cases = cases,
    pop   = pop,
    w     = w,
    nsim  = 199
  )
}

```

Now this is the analysis only for the northen states

```{r}
Hosp_pop_north_2010 <- Hosp_pop_north %>%
  filter(Year == 2010)
table(Hosp_pop_north_2010$state_code)

coords <- as.matrix(Hosp_pop_north_2010[, c("lon.x", "lat.x")])
cases  <- Hosp_pop_north_2010$num_of_total
pop    <- Hosp_pop_north_2010$POB_TOTAL

scan_out_north <- scan.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  nsim   = 499,
  alpha  = 0.05,
  ubpop  = 0.25,     # cluster can include up to 25% of total pop
  longlat = FALSE,
  type   = "poisson"
)

scan_out_north
cl_scan <- clusters(scan_out_north)

Hosp_pop_north_2010$scan_cluster <- "None"
Hosp_pop_north_2010$scan_cluster[cl_scan$location$idx] <- "High hospital density"
library(ggplot2)



```

```{r}
library(sf)
library(rnaturalearth)
library(ggplot2)

mex_states <- st_transform(mex_shp, 4326)  # make sure it's lon/lat
mex_north_sf <- mex_states %>%
  filter(CVE_ENT %in% c("02", "05", "08", "19", "26", "28"))


ggplot() +
  geom_sf(
    data = mex_north_sf,
    fill = "white",
    color = "black",
    linewidth = 0.4
  ) +
  geom_point(
    data = Hosp_pop_north_2010,
    aes(x = lon.x, y = lat.x, color = scan_cluster),
    size = 1.5,
    alpha = 0.9
  ) +
  scale_color_manual(values = c(
    "None" = "gray70",
    "High hospital density" = "#d73027"
  )) +
  coord_sf() +
  labs(
    title = "Hospital Supply Clusters in Northern Border States (2010)",
    subtitle = "Kulldorff spatial scan with state boundaries",
    x = "Longitude",
    y = "Latitude",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )


```

```{r}
# Make sure CLAVE is 5-character string in your data
Hosp_pop_north_2010 <- Hosp_pop_north_2010 %>%
  mutate(CLAVE = sprintf("%05d", as.integer(CLAVE)))

# Create full municipal ID in shapefile: SS + MMM
mex_north_sf <- mex_north_sf %>%
  mutate(
    CLAVE = paste0(
      sprintf("%02d", as.integer(CVE_ENT)),
      sprintf("%03d", as.integer(CVE_MUN))
    )
  )

mex_north_joined <- mex_north_sf %>%
  left_join(
    Hosp_pop_north_2010,
    by = "CLAVE"
  )

library(ggplot2)

ggplot(mex_north_joined) +
  geom_sf(aes(fill = num_of_total), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", na.value = "white") +
  labs(
    title = "Hospital Counts by Municipality (Northern Border States, 2010)",
    fill = "Hospitals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

```{r}
ggplot(mex_north_joined) +
  geom_sf(aes(fill = POB_TOTAL), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "white",
                       labels = scales::comma) +
  labs(
    title = "Population by Municipality (Northern Border States, 2010)",
    fill = "Population"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

```{r}
mex_north_joined <- mex_north_joined %>%
  mutate(
    hosp_per_10k = (num_of_total / POB_TOTAL) * 10000
  )

ggplot(mex_north_joined) +
  geom_sf(aes(fill = hosp_per_10k), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "cividis", na.value = "white") +
  labs(
    title = "Hospitals per 10,000 Residents (Northern Border States, 2010)",
    fill = "Hosp / 10k"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

```{r}
Hosp_pop_2010 <- Hosp_pop_xy %>%
  filter(Year == 2010) %>%
  mutate(
    CLAVE = sprintf("%05d", as.integer(CLAVE)),
    hosp_per_10k = (num_of_total / POB_TOTAL) * 10000
  )
mex_all_sf <- mex_shp %>%
  st_transform(4326) %>%
  mutate(
    CVE_ENT = sprintf("%02d", as.integer(CVE_ENT)),
    CVE_MUN = sprintf("%03d", as.integer(CVE_MUN)),
    CLAVE   = paste0(CVE_ENT, CVE_MUN)
  )
mex_all_joined <- mex_all_sf %>%
  left_join(
    Hosp_pop_2010,
    by = "CLAVE"
  )
library(ggplot2)

ggplot(mex_all_joined) +
  geom_sf(aes(fill = hosp_per_10k), color = "gray40", linewidth = 0.05) +
  scale_fill_viridis_c(option = "cividis", na.value = "white") +
  labs(
    title = "Hospitals per 10,000 Residents (Mexico, 2010)",
    fill = "Hosp / 10k"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid  = element_blank()
  )

```
