---
title: "Hospitals Mexico Project"
author: "Kevin Ruiz"
format: html
editor: visual
---

```{r}
library(readr)
library(sf)
library(readxl)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(smerc)
```

# Loading In Data and Pre-Processing

```{r}
Merged_CLUES_2016 <- read_csv("Merged_CLUES_2016.csv")
head(Merged_CLUES_2016)
unzip("mx_shp.zip", exdir = "mx_shp")
mex_shp <- st_read("mx_shp/mx.shp")
mex_shp2 <- sf::st_read("mx_shp/mx.shp")

Merged_CLUES_2016_clean <- Merged_CLUES_2016 |>
  dplyr::filter(!is.na(longitude), !is.na(latitud)) 

pts_sf <- st_as_sf(Merged_CLUES_2016_clean, coords = c("longitude", "latitud"),  crs = 4326)

plot(st_geometry(mex_shp), col = "white", border = "gray")
plot(st_geometry(pts_sf), col = "red", pch = 5, add = TRUE, cex = .5, size =1)

```

Some exploratory graphs showing where we have a lot of our data, for example in the state of chihuahua we see a lot of our health facilities somewhat clustered in the regions near all of the big cities with little to none rural ones.

Now for us to actually do some of the analysis I went ahead and aggregated the data by state municipality code and year. This can help us use spatial statistic methods used in regional count data which helps simplify our analysis.

```{r}
#lets load in our dataset with the aggregation done. 

data <- read_csv("Num_of_Clinics.csv")
pop <- read_excel("00_Republica_mexicana/1_Grupo_Quinq_00_RM.xlsx") 

```

With our data loaded in we can start by doing some basic data exploration. First, however, lets get the population by municipality for each year.

```{r}

pop_total_by_gender <- pop %>%
  select(CLAVE, POB_TOTAL, ANO) %>%
  rename(Year = ANO) %>%
  filter(Year == 2025)

#Getting our actual poulation total

pop_total <- pop_total_by_gender %>%
  group_by(CLAVE, Year) %>%
  summarise(
    POB_TOTAL = sum(POB_TOTAL, na.rm = T),
    .groups = "drop"
  )

# Doing a rename so that we can combine the data
data_clave <- data %>%
  rename(CLAVE = State_Municipality)

# Number of hopsitals plus a column for our population total for that year. 
Hosp_pop <- data_clave %>%
  left_join(pop_total, by = c("CLAVE"))
```

```{r}
## Loading in our shape file for the municipaliteis goetten from government website

mex_shp <- st_read("mun21gw/mun21gw.shp")

# Using SF to get centroids for the shapes
mun_centroids <- st_centroid(mex_shp)

# Getting Centroid coords
coords <- st_coordinates(mun_centroids)

# Making the CLAVE match up as well as extracting our coordinates
mun_centroids <- mun_centroids %>%
  mutate(
    CVE_ENT = as.numeric(CVE_ENT),
    CVE_MUN = as.numeric(CVE_MUN),
    CLAVE = CVE_ENT * 1000 + CVE_MUN,
    lon = coords[,1],
    lat = coords[,2]
  )

# Having our coords 
names(mun_centroids)
mun_xy <- mun_centroids %>%
  st_drop_geometry() %>%
  select(CLAVE, lon, lat)

# This is all of the municiplaites
Hosp_pop_xy <- Hosp_pop %>%
  left_join(mun_xy, by = "CLAVE") %>%
  filter(!is.na(POB_TOTAL))


# With this we are only looking at the municiplaites in the northern part of mexico
north_states <- c("02", "05", "08", "19", "26", "28")

Hosp_pop_north <- Hosp_pop_xy %>%
  left_join(mun_xy, by = "CLAVE") %>%
  mutate(CLAVE = sprintf("%05d", as.integer(CLAVE))) %>%
  mutate(state_code = substr(CLAVE, 1, 2)) %>%
  filter(state_code %in% north_states) 



```

# National Analysis

## Setting Our Data sets

```{r}
coords <- as.matrix(Hosp_pop_xy[, c("lon", "lat")])
cases  <- Hosp_pop_xy$num_of_total
pop    <- Hosp_pop_xy$POB_TOTAL
ex     <- sum(cases) / sum(pop) * pop  
```

## Besag and Newell Method

This is the Besag Newell Method on the whole population of Mexico.

Ho: The most compact window regarding population with at least c\* hospitals is not significantly more compact than what is expected under the constant risk hypothesis

Ha: The most compact window regarding population with at least c\* hospitals is significantly more compact than what is expected under the constant risk hypothesis

```{r}
cstar <- 5

bn_out <- bn.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  cstar  = cstar,
  alpha  = 0.05,
  longlat = FALSE,
  modified = FALSE
)

bn_out  


cl_bn <- clusters(bn_out)

cl_bn$location$idx

Hosp_pop_xy$bn_cluster <- "None"

Hosp_pop_xy$bn_cluster[cl_bn$location$idx] <- "High hospital density"
```

Lets plot the B_N Method

```{r}
ggplot(Hosp_pop_xy, aes(x = lon, y = lat)) +
  geom_point(aes(color = bn_cluster), size = 1.4, alpha = 0.85) +
  scale_color_viridis_d(
    option = "C",
    name = "BN Cluster"
  ) +
  coord_fixed() +
  labs(
    title = "Besag–Newell Cluster of Hospital Supply (2010)",
    subtitle = paste0("c* = ", cstar, " hospitals"),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
```

## Data Exploration Per Capita Hospital and Diff

Finding the average per capita hospital and the difference

```{r}
Hosp_pop_xy <- Hosp_pop_xy |>
  mutate(
    hosp_per_10k = (num_of_total / POB_TOTAL) * 10000
  )

avg_rate <- mean(Hosp_pop_xy$hosp_per_10k, na.rm = TRUE)

Hosp_pop_xy <- Hosp_pop_xy |>
  mutate(
    deviation_from_avg_rate = hosp_per_10k - avg_rate
  )
mex_shp <- mex_shp %>%
  mutate(
    CVE_ENT_num = as.numeric(as.character(CVE_ENT)),
    CVE_MUN_num = as.numeric(as.character(CVE_MUN)),
    CLAVE = CVE_ENT_num * 1000 + CVE_MUN_num)

mex_map_dev <- mex_shp |>
  left_join(Hosp_pop_xy, by = "CLAVE")


```

The difference plotted in ggplot

```{r}
ggplot(mex_map_dev) +
  geom_sf(aes(fill = deviation_from_avg_rate), color = NA) +
  scale_fill_viridis_c(
    option = "D",   # diverging viridis
    direction = 1,
    name = "Deviation\n(per 10k)"
  ) +
  labs(
    title = "Hospital Access Deviation from National Average (2010)",
    subtitle = "Hospitals per 10,000 − National Mean"
  ) +
  theme_minimal()


```

## Cluster Evaluation Permutation Procedure (CEEP)

Ho: There is no window with n\* persons that has significantly more clinics than what is expected under the CRH

Ha: There is at least one window with n\* persons that has significantly more clinics than what is expected under the CRH

```{r}

N <- sum(pop, na.rm = TRUE)

nstar_vec <- c(0.01, 0.03, 0.05) * N
nstar_vec


cepp_list_01 <-cepp.test(
    coords = coords,
    cases  = cases,
    pop    = pop,
    nstar  = nstar_vec[1],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )
cepp_list_03 <-cepp.test(
    coords = coords,
    cases  = cases,
    pop    = pop,
    nstar  = nstar_vec[2],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )
cepp_list_05 <-cepp.test(
    coords = coords,
    cases  = cases,
    pop    = pop,
    nstar  = nstar_vec[3],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )



```

Lets go ahead and plot the CEEP method

```{r}

plot(sf::st_geometry(mex_shp), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_01))
legend("topright", legend = c("n* = 1333611"))

plot(sf::st_geometry(mex_shp), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_03))
legend("topright", legend = c("n* = 4000833"))

plot(sf::st_geometry(mex_shp), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_05))
legend("topright", legend = c("n* = 6668055"))


```

## Scan Test

```{r}
e = sum(cases)/sum(pop) * pop
scan_out <- scan.test(
  coords = coords,
  cases  = cases,
  pop    = pop,
  nsim   = 499,
  alpha  = 0.05,
  ex = e,
  longlat = T
)

k <- length(scan_out$clusters)
mycol <- grDevices::hcl.colors(k)

plot(
  sf::st_geometry(mex_shp),
  border = "grey60",
  axes   = TRUE,
  col    = color.clusters(scan_out , col = mycol)
)

legend(
  "topright",
  legend = paste("Cluster", seq_len(k)),
  lwd = 8,
  col = mycol
)
```

## Tango's Test

```{r}
coords <- as.matrix(df_year[, c("x", "y")])
w      <- dweights(coords, kappa = 1)  # distance-based weights

tango_out <- tango.test(
  cases = cases,
  pop   = pop,
  w     = w,
  nsim  = 499    # Monte Carlo p-value
)

tango_out
plot(tango_out)

```

```{r}
years <- sort(unique(Hosp_pop_xy$Year))

results_scan  <- list()
results_tango <- list()

for (yy in years) {
  this <- Hosp_pop_xy %>% filter(Year == yy)

  coords <- as.matrix(this[, c("lon", "lat")])
  cases  <- this$num_of_total
  pop    <- this$POB_TOTAL

  # scan
  results_scan[[as.character(yy)]] <- scan.test(
    coords = coords,
    cases  = cases,
    pop    = pop,
    nsim   = 199,
    alpha  = 0.05,
    ubpop  = 0.5
  )

  # tango
  w <- dweights(coords, kappa = 1)
  results_tango[[as.character(yy)]] <- tango.test(
    cases = cases,
    pop   = pop,
    w     = w,
    nsim  = 199
  )
}

```

# Northen States Analysis

## Northen States Plots

```{r}
states_names <- c("Baja California", "Chiapas", "Chihuahua", 
                  "Nuevo León", "Sonora", "Tamaulipas")

states_codes <- c(2,5,8,19,26,28)

CVE_ENT_North <- c("02", "05", "08", "19", "26", "28")

state_sf <- mex_shp |> 
  filter(CVE_ENT %in% CVE_ENT_North)

states_pts <- pts_sf |> 
  filter(state_code %in% states_codes)

plot(st_geometry(state_sf), col = "white", border = "gray")
plot(st_geometry(states_pts), col = "red", add = TRUE, cex = 0.5)
```

```{r}

# Make sure CLAVE is 5-character string in your data
Hosp_pop_north <- Hosp_pop_north %>%
  mutate(CLAVE = sprintf("%05d", as.integer(CLAVE)))

# Create full municipal ID in shapefile: SS + MMM
mex_north_sf <- state_sf %>%
  mutate(
    CLAVE = paste0(
      sprintf("%02d", as.integer(CVE_ENT)),
      sprintf("%03d", as.integer(CVE_MUN))
    )
  )
state_sf <- state_sf %>% mutate(CLAVE = as.numeric(CLAVE))
Hosp_pop_north <- Hosp_pop_north %>% mutate(CLAVE = as.numeric(CLAVE))

mex_north_joined <- state_sf %>%
  left_join(
    Hosp_pop_north,
    by = "CLAVE"
  )



ggplot(mex_north_joined) +
  geom_sf(aes(fill = num_of_total), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", na.value = "white") +
  labs(
    title = "Hospital Counts by Municipality (Northern Border States, 2010)",
    fill = "Hospitals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

```{r}

ggplot(mex_north_joined) +
  geom_sf(aes(fill = POB_TOTAL), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "white",
                       labels = scales::comma) +
  labs(
    title = "Population by Municipality (Northern Border States, 2010)",
    fill = "Population"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```

```{r}
mex_north_joined <- mex_north_joined %>%
  mutate(
    hosp_per_10k = (num_of_total / POB_TOTAL) * 10000
  )

ggplot(mex_north_joined) +
  geom_sf(aes(fill = hosp_per_10k), color = "gray40", linewidth = 0.1) +
  scale_fill_viridis_c(option = "cividis", na.value = "white") +
  labs(
    title = "Hospitals per 10,000 Residents (Northern Border States, 2025)",
    fill = "Hosp / 10k"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```

## Setting Our Data sets Northen States

Let's first create our typical cords, cases, population, and our expected risk under CRH

```{r}
coords_n <- as.matrix(Hosp_pop_north[, c("lon.x", "lat.x")])
cases_n  <- Hosp_pop_north$num_of_total
pop_n    <- Hosp_pop_north$POB_TOTAL
#CRH expected value
ex_n     <- sum(cases) / sum(pop) * pop  
```

## Besag and Newell Approach Northern States

Ho: The most compact window regarding population with at least c\* hospitals in the border U.S. Mexico is not significantly more compact than what is expected under the constant risk hypothesis

Ha: The most compact window regarding population with at least c\* hospitals in the border U.S. Mexico is significantly more compact than what is expected under the constant risk hypothesis

```{r}

summary(cases_n)
quantile(cases_n, probs = c(0.5, 0.75, 0.9))
# I choose mean, 50th, 75th, 90th, 
cstars_n <- c(10.5,  28, 33.32, 68.3)

bn_out_n_10_5 <- bn.test(
  coords = coords_n,
  cases  = cases_n,
  pop    = pop_n,
  cstar  = cstars_n[1],
  alpha  = 0.05,
  longlat = T,
  modified = FALSE
)

bn_out_n_28 <- bn.test(
  coords = coords_n,
  cases  = cases_n,
  pop    = pop_n,
  cstar  = cstars_n[2],
  alpha  = 0.05,
  longlat = T,
  modified = FALSE
)

bn_out_n_33_32 <- bn.test(
  coords = coords_n,
  cases  = cases_n,
  pop    = pop_n,
  cstar  = cstars_n[3],
  alpha  = 0.05,
  longlat = T,
  modified = FALSE
)

bn_out_n_68_3 <- bn.test(
  coords = coords_n,
  cases  = cases_n,
  pop    = pop_n,
  cstar  = cstars_n[4],
  alpha  = 0.05,
  longlat = T,
  modified = FALSE
)

```

Now lets go ahead an plot

```{r}
mycol = brewer.pal(4, "Dark2")

# default is white (no clustering)
ncol = rep("white", nrow(mex_north_joined))
# cstar = 10.5
ncol[bn_out_n_10_5$clusters[[1]]$locids] = mycol[2]

# cstar = 28
ncol[bn_out_n_28$clusters[[1]]$locids] = mycol[1]

# cstar = 33.32
ncol[bn_out_n_33_32$clusters[[1]]$locids] = mycol[4]

# cstar = 68.3
ncol[bn_out_n_68_3$clusters[[1]]$locids] = mycol[3]

plot(sf::st_geometry(mex_north_joined),
     border = "grey60",
     axes   = TRUE,
     col    = ncol)

legend("topright",
       legend = c("Cluster c* = 28",
                  "Cluster c* = 10.5",
                  "Cluster c* = 68.3",
                  "Cluster c* = 33.32"),
       lwd = 10,
       col = mycol)
bn_out_n_10_5$clusters[[1]][c("cases", "population", "pvalue")]
bn_out_n_28$clusters[[1]][c("cases", "population", "pvalue")]
bn_out_n_33_32$clusters[[1]][c("cases", "population", "pvalue")]
bn_out_n_68_3$clusters[[1]][c("cases", "population", "pvalue")]

```

## Cluster Evaluation Permutation Procedure Northern States

Ho: There is no window with n\* persons that has significantly more clinics in the U.S. Mexico border states than what is expected under the CRH

Ha: There is at least one window with n\* persons that has significantly more clinicsin the U.S. Mexico border states than what is expected under the CRH

Now lets actually run the CEPP Model

```{r}
N <- sum(pop_n, na.rm = TRUE)

nstar_vec_n <- c(0.01, 0.03, 0.05) * N
nstar_vec_n


cepp_list_01_n <-cepp.test(
    coords = coords_n,
    cases  = cases_n,
    pop    = pop_n,
    nstar  = nstar_vec_n[1],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )
cepp_list_03_n <-cepp.test(
    coords = coords_n,
    cases  = cases_n,
    pop    = pop_n,
    nstar  = nstar_vec_n[2],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )
cepp_list_05_n <-cepp.test(
    coords = coords_n,
    cases  = cases_n,
    pop    = pop_n,
    nstar  = nstar_vec_n[3],
    nsim   = 499,
    alpha  = 0.05,
    longlat = T
  )


```

Lets go ahead and plot our tests

```{r}

plot(sf::st_geometry(mex_north_joined), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_01_n))
legend("topright", legend = c("n* = 248773.1"))

plot(sf::st_geometry(mex_north_joined), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_03_n))
legend("topright", legend = c("n* = 746319.3"))

plot(sf::st_geometry(mex_north_joined), border = "white", axes = TRUE,
     col = color.clusters(cepp_list_05_n))
legend("topright", legend = c("n* = 1243865.5"))

```

## Scan Test (Kulldorff Scan)

```{r}
e = sum(cases_n)/sum(pop_n) * pop_n
scan_out_north <- scan.test(
  coords = coords_n,
  cases  = cases_n,
  pop    = pop_n,
  nsim   = 499,
  alpha  = 0.05,
  ex = e,
  longlat = T
)

k <- length(scan_out_north$clusters)
mycol <- grDevices::hcl.colors(k)

plot(
  sf::st_geometry(mex_north_joined),
  border = "grey60",
  axes   = TRUE,
  col    = color.clusters(scan_out_north, col = mycol)
)

legend(
  "topright",
  legend = paste("Cluster", seq_len(k)),
  lwd = 8,
  col = mycol
)
```

Now lets extract the exact municipalities where they're are clusters

```{r}

idx1 <- scan_out_north$clusters[[1]]
idx1 <- idx1$locids


mex_north_joined$NOM_MUN[idx1]
```

```{r}
cluster_rows_df <- mex_north_joined[idx1, ] |>
  sf::st_drop_geometry() |>
  mutate(
    border_us = ifelse(NOM_MUN %in% c("Manuel Benavides", "Matamoros", "Nogales", "Janos"),
                       "Yes", "No")
  ) |>
  select(NOM_MUN, CVE_ENT, border_us)
```

```{r}
idx1 <- scan_sum$iidx[[1]]

cluster_col <- rep("white", nrow(mex_north_joined))
cluster_col[idx1] <- "red"

plot(sf::st_geometry(mex_north_joined),
     border = "grey85",      # soften background borders
     axes   = TRUE,
     col    = cluster_col)

# Thicken ONLY the cluster border
plot(sf::st_geometry(mex_north_joined)[idx1],
     add = TRUE,
     border = "black",
     lwd = 2.5)

legend("topright",
       legend = "Primary Scan Cluster",
       lwd = 10,
       col = "red")

```

```{r}
Hosp_pop_2010 <- Hosp_pop_xy %>%
  filter(Year == 2025) %>%
  mutate(
    CLAVE = sprintf("%05d", as.integer(CLAVE)),
    hosp_per_10k = (num_of_total / POB_TOTAL) * 10000
  )
mex_all_sf <- mex_shp %>%
  st_transform(4326) %>%
  mutate(
    CVE_ENT = sprintf("%02d", as.integer(CVE_ENT)),
    CVE_MUN = sprintf("%03d", as.integer(CVE_MUN)),
    CLAVE   = paste0(CVE_ENT, CVE_MUN)
  )
mex_all_joined <- mex_all_sf %>%
  left_join(
    Hosp_pop_2010,
    by = "CLAVE"
  )
library(ggplot2)

ggplot(mex_all_joined) +
  geom_sf(aes(fill = hosp_per_10k), color = "gray40", linewidth = 0.05) +
  scale_fill_viridis_c(option = "cividis", na.value = "white") +
  labs(
    title = "Hospitals per 10,000 Residents (Mexico, 2010)",
    fill = "Hosp / 10k"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid  = element_blank()
  )

```

## Moran's I with Local Rates

```{r}

hospnb <- poly2nb(mex_north_joined)
w = nb2mat(hospnb, style = "B")
# see ?nb2listw for more options
mex_north_joined$num_of_total

```

```{r}
smerc::morancr.test(cases = cases_n, pop = pop_n, w = w, ex = ex_n)
```
